### 2022-10-19 manuscript.Rmd
Fix some typos and formatting issues

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC).

@@

### 1.2.1	Statistical Power to Detect a Specific Effect of Interest
Central to many studies that employ SEM is the question of whether a specific parameter in a model (e.g., strength of a specific correlation or the slope parameter in a latent regression model) is different from a given value [e.g. @canada_2021]. Imagine that a researcher aims to test whether the correlation between two latent constructs significantly differs from zero. To this end, they would compare the model-fit between one model in which the correlation parameter is freely estimated and another model in which it is fixed to zero [@satorra_1985]. The corresponding null hypothesis in this case states that the parameter restrictions hold in the population. The difference in model fit will follow a $\chi^2$-distribution with degrees-of-freedom (*df*) equal to the difference of freely estimated parameters between the two models if the null hypothesis is true [@neale_2000, in this example *df* = 1, because only the correlation between the two latent factors is fixed in the restricted model]. If restricting the parameter of interest results in a significantly poorer model fit, this would suggest that the respective parameter indeed significantly differs from zero.
Following the logic outlined above, researchers can determine the statistical power for detecting a target effect in an SEM framework before conducting the study. A key aspect for this a-priori power calculation hinges on translating the differences in a specific parameter estimate into an effect size. Such translation can be achieved by investigating the discrepancy between the model-implied variance-covariance matrices associated with (1) the population model including the “true” parameter values and (2) the hypothesized model. The discrepancy between both matrices is quantified based on a specific fit function [for details on fit functions see @bollen_1989]. Recently, several user friendly tools have emerged that can guide modelers in quantifying effect sizes of a target effect to estimate statistical power in SEM [e.g. the R packages semPower @jobst_2021; or power4SEM @jak_2020 for analytical approaches; and the Shiny app pwrSEM @wang_2021; the interactive study planner tool LIFESPAN @brandmaier_2015;, or the R package simsem  @pornprasertmanit_2021 for simulation-based approaches]. As the investigation of a target effect practically translates into the comparison of nested models, researchers can in these cases analytically determine the statistical power and use this information for decisions on required sample size [@satorra_1985].

@@

## 2.1 How does a simulation work: Power-estimations for the comparison of non-nested SEMs
To investigate statistical power for deciding between competing theoretical non-nested models, we suggest a Monte Carlo simulation-based approach [@metropolis_1953]. Monte Carlo studies represent a computer-intensive simulation, that allows to approximate statistical power for a given study design [@muthen_2002]. While Monte Carlo simulations can be implemented for a variety of research questions, the flexibility of the approach makes them especially well-suited for obtaining a-priori power estimates for the comparison of non-nested SEMs. Furthermore, analytical approaches are tied to assumptions (e.g., no missing data, multivariate normality), whereas simulation-based approaches allow for arbitrary data generating processes. In Monte Carlo studies, synthetic data is repeatedly generated with a set of different hypothesized parameter values and analyzed across all samples. Summary statistics from the entirety of the simulated datasets are then used to draw conclusions [@muthen_2002; @paxton_2001]. The general procedure of Monte Carlo simulations follows a common core structure: (i) generate – in which multiple synthetic datasets are generated based on the hypothesized models, (ii) analyze – in which the synthetic data are analyzed for every iteration of the simulation, (iii) summarize – in which the results are pooled over all simulation iterations. This general procedure is applicable to a variety of scenarios. When comparing competing SEMs, it is necessary to specify the competing theoretical models, and decide on the simulation design, that is which model parameters of interest should be modified across the simulations. Finally, researchers need to evaluate the results from the simulation in order to derive practical implications from the simulation (see Figure 2). While these steps can be manually coded in a statistical research software, such as R, the implementation of a simulation can become increasingly complex and error-prone with increasing number of conditions that are being simulated. A useful tool which can guide a novice simulators in setting up a Monte Carlo simulation is the R package simDesign [@chalmers_2020], which facilitates the implementation of the internal logic of generate – analyze – simulate and can accommodate various research designs. All simulation steps in this study where therefore implemented within the simDesign package.

@@

### 2.1.1	Step 1: Specifying Competing SEMs
The first step of the simulation is translating the theoretical assumptions about latent constructs of interest and the manifest variables into an SEM. This step entails a) specifying the number of latent factors (how many theoretical constructs are thought to influence the observed behavior in the manifest variables?), and b) the allocation of each manifest variable to (at least one) latent variable (which of the manifest variables captures the respective latent construct?). In R, the structure of a SEM can be easily implemented using lavaan syntax [@rosseel_2012]. Typically, useful operators for specifying a model entail factor loadings (=~), (co )variances (~~), and means or intercepts (~1). The tri-partite model (Figure 1, Model 3) could be specified in lavaan syntax as the following:

@@

Estimating statistical power for model comparisons requires that researchers specify each of the competing models and all aspects in which they differ from one another (e.g., in number of latent factors, allocations of manifest variables; see supplementary R code on github for specifications of Models 1 and 2).

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as a for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as b to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). The population model is defined by a model string, which specifies three latent factors (ps, pc, gen), measured by the 13 manifest indicators (task1 – task13). The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The covariance between the latent factors is set to the respective value from the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package and stored as dat. For the sake of clarity, all loadings and co-variances are spelled out separately (but see supplementary R code on github for rewriting the same model in a more concise syntax).

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters. We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package and extract different measures of model fit using the helper function get_fitmeasures() (see funs.R in the supplementary material on github), namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

### 2.1.5	Step 5: Summarize
In step 5, we define how the results from the analyze step should be summarized. To this end, we extract how often the model used to generate the data (step 2) was actually recovered as the best fitting model. We implemented the following decision rules for identifying the best fitting model: (i) We define cut-off criteria for an adequate model fit according to established rules as CFI > .95 and RMSEA < 0.06 [@hu_1999]. (ii) Among the models that meet the cut-off criteria, we select the best fitting model for each data set based on the relative model fit (lowest BIC value). The goal of (i) and (ii) is to check whether the ground truth model that generated the data can be identified as the best fitting model in a given simulation run. For cases in which a given dataset does not converge for one of the models, we consider the outcome of these iterations as failing to recover the correct model. We summarize the proportion of successful model recovery for each combination of model parameters in the simulation separately. The summarize function that concatenates the results across simulation iterations for the example on the structure of memory in childhood could thus look like this:
```{r summarize, echo = TRUE}
```
First, the function restructures the results from the analyze function to obtain a tidy data frame with the fit measures for each model in a separate row. This data is then summarized for each iteration. The function identifies a) the ground truth model, that generated the data in this simulation iteration (sim_model) and b) the model which produced the best model fit (best). Here, the best model is defined as “0” if one of the three theoretical models did not converge, or if none of the models met the cut-off criteria as defined above. Otherwise, the best model is selected among the ones that met the cut-off criteria based on the relative model-fit.  The function returns the argument best_3, indicating how often across all iterations Model 3 was identified as the best fitting model, and best_sim, indicating how often the model that simulated the data was identified as the best fitting model (note that in the current analysis, both values are identical. Best_sim can however be useful to investigate Type-I error, see section 3.3).

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for recovering the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 10,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on github"}

@@

The results of this analysis reveal that even for low levels of loadings strength (< 0.7), the probability of erroneously picking Model 3 as the best fitting model is reasonably low (< 7%). Likewise, this probability remains reasonably low for different levels of inter-factor correlations for all sample sizes (< 5%). Thus, in this specific model comparison Type-I error probability does not appear to be the main aspect driving sample size requirements for an informative research design. 
## 3.4	Deriving a Sample Estimate From the Simulation Results
To derive a joint sample size estimate from the different parameters in the simulation, it is crucial to reiterate the theoretical assumptions on the strength of all parameters investigated in the simulation. In our illustration, it is reasonable to assume that the latent factors in the tri-partite model (Model 3, Figure 1) are only moderately correlated (below 0.7). Therefore, our model comparison requires a sample size of *N* = 120 to reach statistical power of 95% to adjudicate between the theoretical models, if the hypothesized model (Model 3) generates the data. We further assume that the manifest variables capture the latent constructs reasonably well — that is, exerting a loading strength of 0.7 or above. The results of the simulation suggest that *N* = 120 would suffice to achieve 95% with regard to the reliability of the manifest factors. In our example, the results of both analyses align, such that a sample size of *N* = 120 suffices to achieve a power of 95% regarding both the separability of the theoretical constructs and the reliability of the indicators. In cases where the derived sample estimate differs between the two analyses we suggest to derive the more conservative estimate of required sample size from the two simulations.
The results from the Type-I error analysis further indicate, that the probability of erroneously choosing an overly complex model is generally low. Therefore, the probability of committing a Type-I error for the sample size derived from the simulations on model separability and measurement reliability (*N* = 120) is negligible.

@@

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as github or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021].

>>>>

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP is a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC).

@@

### 1.2.1	Statistical Power to Detect a Specific Effect of Interest
Central to many studies that employ SEM is the question of whether a specific parameter in a model (e.g., strength of a specific correlation or the slope parameter in a latent regression model) is different from a given value [e.g. @canada_2021]. Imagine that a researcher aims to test whether the correlation between two latent constructs significantly differs from zero. To this end, they would compare the model fit between one model in which the correlation parameter is freely estimated and another model in which it is fixed to zero [@satorra_1985]. The corresponding null hypothesis in this case states that the parameter restrictions hold in the population. The difference in model fit will follow a $\chi^2$-distribution with degrees-of-freedom (*df*) equal to the difference of freely estimated parameters between the two models if the null hypothesis is true [@neale_2000, in this example *df* = 1, because only the correlation between the two latent factors is fixed in the restricted model]. If restricting the parameter of interest results in a significantly poorer model fit, this would suggest that the respective parameter indeed significantly differs from zero.
Following the logic outlined above, researchers can determine the statistical power for detecting a target effect in an SEM framework before conducting the study. A key aspect for this a-priori power calculation hinges on translating the differences in a specific parameter estimate into an effect size. Such translation can be achieved by investigating the discrepancy between the model-implied variance-covariance matrices associated with (1) the population model including the “true” parameter values and (2) the hypothesized model. The discrepancy between both matrices is quantified based on a specific fit function [for details on fit functions see @bollen_1989]. Recently, several user friendly tools have emerged that can guide modelers in quantifying effect sizes of a target effect to estimate statistical power in SEM [e.g. the R packages semPower @jobst_2021; or power4SEM @jak_2020 for analytical approaches; and the Shiny app pwrSEM @wang_2021; the interactive study planner tool LIFESPAN @brandmaier_2015; or the R package simsem  @pornprasertmanit_2021 for simulation-based approaches]. As the investigation of a target effect practically translates into the comparison of nested models, researchers can in these cases analytically determine the statistical power and use this information for decisions on required sample size [@satorra_1985].

@@

## 2.1 How does a simulation work: Power-estimations for the comparison of non-nested SEMs
To investigate statistical power for deciding between competing theoretical non-nested models, we suggest a Monte Carlo simulation-based approach [@metropolis_1953]. Monte Carlo studies represent a computer-intensive simulation, that allows to approximate statistical power for a given study design [@muthen_2002]. While Monte Carlo simulations can be implemented for a variety of research questions, the flexibility of the approach makes them especially well-suited for obtaining a-priori power estimates for the comparison of non-nested SEMs. Furthermore, analytical approaches are tied to assumptions (e.g., no missing data, multivariate normality), whereas simulation-based approaches allow for arbitrary data generating processes. In Monte Carlo studies, synthetic data is repeatedly generated with a set of different hypothesized parameter values and analyzed across all samples. Summary statistics from the entirety of the simulated datasets are then used to draw conclusions [@muthen_2002; @paxton_2001]. The general procedure of Monte Carlo simulations follows a common core structure: (i) generate – in which multiple synthetic datasets are generated based on the hypothesized models, (ii) analyze – in which the synthetic data are analyzed for every iteration of the simulation, (iii) summarize – in which the results are pooled over all simulation iterations. This general procedure is applicable to a variety of scenarios. When comparing competing SEMs, it is necessary to specify the competing theoretical models, and decide on the simulation design, that is which model parameters of interest should be modified across the simulations. Finally, researchers need to evaluate the results from the simulation in order to derive practical implications from the simulation (see Figure 2). While these steps can be manually coded in a statistical research software, such as R, the implementation of a simulation can become increasingly complex and error-prone with increasing number of conditions that are being simulated. A useful tool which can guide novice simulators in setting up a Monte Carlo simulation is the R package simDesign [@chalmers_2020], which facilitates the implementation of the internal logic of generate – analyze – simulate and can accommodate various research designs. All simulation steps in this study where therefore implemented within the simDesign package.

@@

### 2.1.1	Step 1: Specifying Competing SEMs
The first step of the simulation is translating the theoretical assumptions about latent constructs of interest and the manifest variables into an SEM. This step entails a) specifying the number of latent factors (how many theoretical constructs are thought to influence the observed behavior in the manifest variables?), and b) the allocation of each manifest variable to (at least one) latent variable (which of the manifest variables captures the respective latent construct?). In R, the structure of a SEM can be easily implemented using lavaan syntax [@rosseel_2012]. Typically, useful operators for specifying a model entail factor loadings (=~), (co-)variances (~~), and means or intercepts (~1). The tri-partite model (Figure 1, Model 3) could be specified in lavaan syntax as the following:

@@

Estimating statistical power for model comparisons requires that researchers specify each of the competing models and all aspects in which they differ from one another (e.g., in number of latent factors, allocations of manifest variables; see supplementary R code on GitHub for specifications of Models 1 and 2).

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as a for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as b to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). The population model is defined by a model string, which specifies three latent factors (ps, pc, gen), measured by the 13 manifest indicators (task1 – task13). The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The covariance between the latent factors is set to the respective value from the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package and stored as dat. For the sake of clarity, all loadings and co-variances are spelled out separately (but see supplementary R code on GitHub for rewriting the same model in a more concise syntax).

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters. We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package and extract different measures of model fit using the helper function get_fitmeasures() (see funs.R in the supplementary material on GitHub), namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

### 2.1.5	Step 5: Summarize
In step 5, we define how the results from the analyze step should be summarized. To this end, we extract how often the model used to generate the data (step 2) was actually recovered as the best fitting model. We implemented the following decision rules for identifying the best fitting model: (i) We define cut-off criteria for an adequate model fit according to established rules as CFI > .95 and RMSEA < 0.06 [@hu_1999]. (ii) Among the models that meet the cut-off criteria, we select the best fitting model for each data set based on the relative model fit (lowest BIC value). The goal of (i) and (ii) is to check whether the ground truth model that generated the data can be identified as the best fitting model in a given simulation run. For cases in which a given dataset does not converge for one of the models, we consider the outcome of these iterations as failing to recover the correct model. We summarize the proportion of successful model recovery for each combination of model parameters in the simulation separately. The summarize function that concatenates the results across simulation iterations for the example on the structure of memory in childhood could thus look like this:
```{r summarize, echo = TRUE}
```
First, the function restructures the results from the analyze function to obtain a tidy data frame with the fit measures for each model in a separate row. This data is then summarized for each iteration. The function identifies a) the ground truth model that generated the data in this simulation iteration (sim_model) and b) the model which produced the best model fit (best). Here, the best model is defined as “0” if one of the three theoretical models did not converge, or if none of the models met the cut-off criteria as defined above. Otherwise, the best model is selected among the ones that met the cut-off criteria based on the relative model-fit.  The function returns the argument best_3, indicating how often across all iterations Model 3 was identified as the best fitting model, and best_sim, indicating how often the model that simulated the data was identified as the best fitting model (note that in the current analysis, both values are identical. best_sim can however be useful to investigate Type-I error, see section 3.3).

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for recovering the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 10,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on GitHub"}

@@

The results of this analysis reveal that even for low levels of loadings strength (< 0.7), the probability of erroneously picking Model 3 as the best fitting model is reasonably low (< 7%). Likewise, this probability remains reasonably low for different levels of inter-factor correlations for all sample sizes (< 5%). Thus, in this specific model comparison Type-I error probability does not appear to be the main aspect driving sample size requirements for an informative research design.

## 3.4	Deriving a Sample Estimate From the Simulation Results
To derive a joint sample size estimate from the different parameters in the simulation, it is crucial to reiterate the theoretical assumptions on the strength of all parameters investigated in the simulation. In our illustration, it is reasonable to assume that the latent factors in the tri-partite model (Model 3, Figure 1) are only moderately correlated (below 0.7). Therefore, our model comparison requires a sample size of *N* = 120 to reach statistical power of 95% to adjudicate between the theoretical models, if the hypothesized model (Model 3) generates the data. We further assume that the manifest variables capture the latent constructs reasonably well — that is, exerting a loading strength of 0.7 or above. The results of the simulation suggest that *N* = 120 would suffice to achieve 95% with regard to the reliability of the manifest factors. In our example, the results of both analyses align, such that a sample size of *N* = 120 suffices to achieve a power of 95% regarding both the separability of the theoretical constructs and the reliability of the indicators. In cases where the derived sample estimate differs between the two analyses we suggest to derive the more conservative estimate of required sample size from the two simulations.
The results from the Type-I error analysis further indicate, that the probability of erroneously choosing an overly complex model is generally low. Therefore, the probability of committing a Type-I error for the sample size derived from the simulations on model separability and measurement reliability (*N* = 120) is negligible.

@@

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as GitHub or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021].
